---
title: "Figures for article"
subtitle: "Modelado espacial de pedodiversidad"
author: "Carlos M. Guío Blanco"
format: html
editor: visual
---

## Figure 1. Response variable

Figure 1. Spatial and statistical representation of the diversity response variable.  
(A) Rao’s quadratic entropy (Q); (B) log-transformed taxonomic diversity per area [log₁₀(Q/A)];  
(C) spatial map of Q; and (D) spatial map of log₁₀(Q/A).

```{r}

# Generic map function
map_resp <- function(df, var, legend_lab, trans = "identity", breaks = waiver(), labels_fn = waiver()) {
  ggplot(df) +
    geom_sf(aes(fill = {{ var }}), colour = NA) +
    scale_fill_gradientn(
      colours  = pal_resp,
      na.value = "white",
      trans    = trans,
      guide    = guide_colourbar(
        direction      = "vertical",
        barwidth       = unit(0.25, "cm"),
        barheight      = unit(3.5, "cm"),
        title.position = "top"
      ),
      name     = legend_lab,
      breaks   = breaks,
      labels   = labels_fn
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.justification = c("right", "top"))
}

# Histograms and maps
p_Q <- ggplot(ucs_rao_sf_4326, aes(x = Q)) +
  geom_histogram(fill = pal_resp[7], bins = 30, colour = "white") +
  labs(x = "Q", y = "Count") +
  theme_minimal(base_size = 12)

p_logQdens <- ggplot(ucs_rao_sf_4326, aes(x = Qdens)) +
  geom_histogram(fill = pal_resp[7], bins = 30, colour = "white") +
  scale_x_log10(breaks = scales::log_breaks(n = 4), labels = scales::label_scientific()) +
  labs(x = expression("log"[10] * "(Q/A)"), y = "Count") +
  theme_minimal(base_size = 12)

p_map_Q <- map_resp(ucs_rao_sf_4326, Q, "Q")

rng_qd     <- range(ucs_rao_sf_4326$Qdens, na.rm = TRUE)
log_breaks <- scales::log_breaks(n = 4)(rng_qd)
p_map_logQdens <- map_resp(ucs_rao_sf_4326, Qdens, "Q/A", trans = "log10", breaks = log_breaks, labels_fn = scales::label_scientific())

p_respuesta_eda <- (p_Q | p_logQdens) / (p_map_Q | p_map_logQdens) + plot_layout(heights = c(1, 3))

ggsave(
  here::here("Figures", "response_eda_en.eps"),
  plot   = p_respuesta_eda,
  width  = 10,
  height = 10,
  dpi    = 350,
  device = cairo_ps
)
markdown
Copiar
Editar


```



## Figure 2. Environmental covariates

Figure 2. Spatial distribution of environmental covariates.  
(A–D) mean values: (A) elevation (m), (B) slope (°), (C) land surface temperature (°C), (D) precipitation (mm);  
(E) geological units; (F–I) coefficient of variation (log-scale) for each covariate.  
Panel (J) left blank for symmetry.


```{r Figure_2}
# ──────────────────────────────────────────────────────────────────────
# Figura: Mapas de media y CV de variables ambientales (mosaico 4x2)
# ──────────────────────────────────────────────────────────────────────

# Cargar patchwork
library(patchwork)

# Paletas
mono_pal <- function(col_hex, n = 100) colorRampPalette(c("#ffffff", col_hex))(n)
pal_dem_cv        <- mono_pal(head(pal_dem,   1))
pal_pendiente_cv  <- mono_pal(tail(pal_slope, 1))
pal_lst_cv        <- mono_pal(tail(pal_lst,   1))
pal_prec_cv       <- mono_pal(tail(pal_prec,  1))

# Función para mapas de magnitud
map_mag <- function(data, var, pal, titulo, leyenda){
  ggplot(data) +
    geom_sf(aes(fill = {{ var }}), colour = NA) +
    scale_fill_gradientn(
      colours = pal, na.value = "white",
      guide   = guide_colourbar(barwidth = unit(0.25,"cm"),
                                barheight = unit(2.5,"cm"),
                                direction = "vertical",
                                title.position = "top"),
      name = leyenda
    ) +
    labs(title = titulo) +
    coord_sf(expand = FALSE) +
    scale_x_continuous(breaks = seq(-79, -70, by = 3)) +
    theme_minimal() +
    theme(legend.justification = c("right","top"),
          plot.title = element_text(size = 10, hjust = .5))
}

# Función para mapas de CV
map_cv <- function(data, var, pal, titulo){
  ggplot(dplyr::filter(data, {{ var }} > 0)) +
    geom_sf(aes(fill = {{ var }}), colour = NA) +
    scale_fill_gradientn(
      colours = pal,
      trans   = "log10",
      breaks  = scales::trans_breaks("log10", function(x) 10^x),
      labels  = scales::label_scientific(digits = 1),
      na.value = "white",
      guide   = guide_colourbar(barwidth = unit(0.25,"cm"),
                                barheight = unit(2.5,"cm"),
                                direction = "vertical",
                                title.position = "top"),
      name    = "CV"
    ) +
    labs(title = titulo) +
    coord_sf(expand = FALSE) +
    scale_x_continuous(breaks = seq(-79, -70, by = 3)) +
    theme_minimal() +
    theme(legend.justification = c("right","top"),
          plot.title = element_text(size = 10, hjust = .5))
}

# Mapas
p_dem_media    <- map_mag(modelo_df, dem_media,       pal_dem,   "Mean elevation (m)", "m")
p_pend_media   <- map_mag(modelo_df, pendiente_media, pal_slope, "Mean slope (°)", "°")
p_lst_media    <- map_mag(modelo_df, lst_media,       pal_lst,   "Mean LST (°C)", "°C")
p_prec_media   <- map_mag(modelo_df, precip_media,    pal_prec,  "Mean precipitation (mm)", "mm")

p_dem_cv       <- map_cv(modelo_df, dem_cv,            pal_dem_cv,       "Elevation CV")
p_pend_cv      <- map_cv(modelo_df, pendiente_cv,      pal_pendiente_cv, "Slope CV")
p_lst_cv       <- map_cv(modelo_df, lst_cv_temporal,   pal_lst_cv,       "LST CV")
p_prec_cv      <- map_cv(modelo_df, precip_cv_temporal,pal_prec_cv,      "Precipitation CV")

# Mosaico
blank <- ggplot() + theme_void()

layout <- area(
  t = c(1,1,1,1,1, 2,2,2,2,2),
  l = c(1,2,3,4,5, 1,2,3,4,5),
  b = c(1,1,1,1,1, 2,2,2,2,2),
  r = c(1,2,3,4,5, 1,2,3,4,5)
)

p_mosaico <- p_dem_media + p_pend_media + p_lst_media + p_prec_media + p_mapa_geo +
             p_dem_cv    + p_pend_cv    + p_lst_cv    + p_prec_cv    + blank +
             plot_layout(design = layout)

# Exportar
ggsave(
  here::here("Figures", "English", "mosaic_covariates_en.png"),
  plot   = p_mosaico,
  width  = 12,
  height = 6,
  dpi    = 350
)


```


Figure 3. Histogram and map of Queen neighborhood degree.  
The histogram shows the frequency of polygons with different numbers of neighbors.  
The map highlights spatial distribution of degree values, helping detect super-nodes or linear artifacts.

```{r Figure_3}


# Histogram
gg_hist_degree_en <- ggplot(
  data.frame(grado = grado_vec) |> dplyr::filter(grado <= 30),
  aes(x = grado)
) +
  geom_histogram(
    binwidth = 1,
    fill     = pal_qualitat[1],
    colour   = "white",
    boundary = 0, closed = "left"
  ) +
  labs(
    x = "Number of neighbours (Queen degree)",
    y = "Polygon count"
  ) +
  theme_minimal()

# Map
p_map_queen_degree_en <- ggplot(modelo_sf_s |> dplyr::filter(grado <= 200)) +
  geom_sf(aes(fill = grado), colour = NA) +
  scale_fill_gradientn(
    colours = pal_continuo,
    guide   = guide_colourbar(
      barwidth = unit(0.25, "cm"),
      barheight = unit(4, "cm"),
      direction = "vertical",
      title.position = "top"
    ),
    name = "Neighbours"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.justification = c("right", "top"))

# Combine
mosaic_degree_en <- p_map_queen_degree_en / gg_hist_degree_en + plot_layout(heights = c(3, 1))

# Export EPS
ggsave(
  here::here("Figures","English", "Figure 3.eps"),
  plot   = mosaic_degree_en,
  width  = 8,
  height = 10,
  dpi    = 350,
  device = cairo_ps
)
```

## Figure 4. Local indicators of spatial association (LISA)

Figure 4. Local Moran’s I tripanel for standardized log₁₀(Q/area):  
(A) Moran scatterplot; (B) LISA cluster map; (C) local Ii statistic.  
These panels visualize clustering patterns and outliers prior to spatial modeling.

```{r Figure_4}
# Prepare data
plot_df <- modelo_sf_s[keep_log, ]
scatter_df <- plot_df |> 
  mutate(z_logQ = as.numeric(scale(log_Qdens)),
         z_lag  = as.numeric(scale(lag_raw))) |> 
  filter(!is.na(cluster_raw))

# Cluster palette
pal_cont <- paletteer::paletteer_c("grDevices::Zissou 1", 100)
col_lisa <- c("High-High" = pal_cont[90],
              "Low-Low"   = pal_cont[10],
              "High-Low"  = pal_cont[70],
              "Low-High"  = pal_cont[30],
              "No sig."   = "grey90")
levels_lisa <- c("No sig.", "Low-Low", "Low-High", "High-Low", "High-High")
scatter_df <- scatter_df |> 
  mutate(cluster_raw = factor(cluster_raw, levels = levels_lisa),
         cluster_sig = ifelse(cluster_raw == "No sig.", "No", "Yes"))

# Global I label
txt_I <- sprintf("Global Moran I = %.4f (p = %.3g)", I_global, p_global)

# Scatterplot
p_scatter <- ggplot(scatter_df,
                    aes(z_logQ, z_lag, colour = cluster_raw, alpha = cluster_sig)) +
  geom_point(size = .8) +
  scale_alpha_manual(values = c("No" = 1, "Yes" = .8), guide = "none") +
  geom_smooth(aes(group = 1), method = "lm", se = FALSE,
              colour = "black", linewidth = .4) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  scale_colour_manual(values = col_lisa, name = "LISA cluster", drop = FALSE) +
  annotate("text", x = -Inf, y =  Inf, hjust = -0.04, vjust = 1.1,
           label = txt_I, size = 3.3, fontface = "italic") +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  labs(x = "Standardized log₁₀(Q / area)", y = "Spatial lag (W · z)") +
  theme_minimal(base_size = 11) +
  theme(legend.justification = c("right", "top"))

# Cluster map
p_clusters <- ggplot(plot_df) +
  geom_sf(aes(fill = cluster_raw), colour = NA) +
  scale_fill_manual(values = col_lisa, name = "LISA cluster",
                    drop = FALSE, na.translate = FALSE) +
  theme_minimal(base_size = 11) +
  theme(legend.justification = c("right", "top"))

# Local Ii map
p_ii <- ggplot(plot_df) +
  geom_sf(aes(fill = Ii_raw), colour = NA) +
  scale_fill_viridis_c(name = "Local Ii", option = "magma", direction = -1,
                       guide = guide_colourbar(
                         barwidth  = unit(0.3, "cm"),
                         barheight = unit(3, "cm"),
                         direction = "vertical",
                         title.position = "top"),
                       na.value = "white") +
  theme_minimal(base_size = 11) +
  theme(legend.justification = c("right", "top"))

# Combine
p_lisa_tripanel <- (p_scatter | p_clusters | p_ii) +
  patchwork::plot_layout(guides = "auto")

# Export EPS
ggsave(
  here::here("Figures", "English", "lisa_tripanel_standardised_en.eps"),
  plot   = p_lisa_tripanel,
  width  = 15,
  height = 5,
  dpi    = 350,
  device = cairo_ps
)
```


## Figure 5. Model performance metrics

Figure 5. Comparison of DIC, WAIC, and logML across spatial models.  
Metrics provide complementary evidence on model fit and complexity.


```{r Figure_5}

# ── Model performance metrics extraction ────────────────────────────────────
tabla_modelos <- tibble::tibble(
  Modelo = c("Full spatial (besag + iid + era)",
             "Spatial without era (besag + iid)",
             "Non-spatial (only era)"),
  DIC   = c(fit_besag_sep$dic$dic,
            fit_besag_noera$dic$dic,
            fit_noesp$dic$dic),
  WAIC  = c(fit_besag_sep$waic$waic,
            fit_besag_noera$waic$waic,
            fit_noesp$waic$waic),
  logML = c(fit_besag_sep$mlik[1],
            fit_besag_noera$mlik[1],
            fit_noesp$mlik[1])
)

tabla_modelos_larga <- tabla_modelos |>
  pivot_longer(cols = c(DIC, WAIC, logML), names_to = "Metric", values_to = "Value") |>
  mutate(
    Metric = factor(Metric, levels = c("DIC", "WAIC", "logML")),
    Modelo = factor(Modelo, levels = c(
      "Full spatial (besag + iid + era)",
      "Spatial without era (besag + iid)",
      "Non-spatial (only era)"
    ))
  )

colores_metricas <- c(
  "DIC"   = "#EF5703FF",
  "WAIC"  = "#E3B710FF",
  "logML" = "#3A9AB2FF"
)

p_metricas <- ggplot(tabla_modelos_larga, aes(x = Value, y = Modelo, fill = Metric)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(
    aes(label = round(Value, 0)),
    position = position_dodge(width = 0.7),
    hjust = ifelse(tabla_modelos_larga$Value > 0, 1.05, -0.05),
    vjust = 0.5,
    size  = 3,
    color = "black"
  ) +
  scale_fill_manual(values = colores_metricas) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    x = "Metric value",
    y = NULL,
    fill = "Metric"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.justification = c("right", "top"))


#Exportar PNG
ggsave(
  filename = here::here("Figures", "English", "Figure 5.png"),
  plot     = p_metricas,
  width    = 7,
  height   = 5,
  dpi      = 350
)


# Exportar EPS
ggsave(
  filename = here::here("Figures", "English","Figure 5.eps"),
  plot     = p_metricas,
  width    = 7,
  height   = 5,
  dpi      = 350,
  device   = cairo_ps
)

p_metricas


```


## Figure 6. Fixed effects coefficients

Figure 6. Posterior summaries for βₖ:  
Left panel shows means and 95% credible intervals for each covariate across models;  
right panel shows the full posterior distributions under the spatial model.


```{r figure_6}


# ── 1. Extract posterior summaries ───────────────────────────────────────────
extract_betas <- function(fit, model_label) {
  fit$summary.fixed |>
    tibble::rownames_to_column("covariate") |>
    dplyr::mutate(model = model_label) |>
    dplyr::select(model, covariate, mean, `0.025quant`, `0.975quant`)
}

betas_comparados <- dplyr::bind_rows(
  extract_betas(fit_besag_sep,   "Full spatial"),
  extract_betas(fit_besag_noera, "Spatial w/o era"),
  extract_betas(fit_noesp,       "Non-spatial")
)

covariate_labels <- c(
  "(Intercept)"               = "Intercept",
  "dem_media_z"              = "Mean elevation (z)",
  "pendiente_media_z"        = "Mean slope (z)",
  "lst_media_z"              = "Mean LST (z)",
  "precip_media_z"           = "Mean precipitation (z)",
  "log_dem_cv_z"             = "Elevation CV (log, z)",
  "log_pendiente_cv_z"       = "Slope CV (log, z)",
  "log_lst_cv_temporal_z"    = "LST CV (log, z)",
  "log_precip_cv_temporal_z" = "Precipitation CV (log, z)"
)

betas_comparados$covariate <- factor(
  betas_comparados$covariate,
  levels = rev(names(covariate_labels))
)
levels(betas_comparados$covariate) <- rev(unname(covariate_labels))

# ── 2. Forest plot ────────────────────────────────────────────────────────────
pal_modelos <- c(
  "Full spatial"     = "#EF5703FF",
  "Spatial w/o era"  = "#E3B710FF",
  "Non-spatial"      = "#3A9AB2FF"
)

p_forest <- ggplot(betas_comparados, aes(
  x = mean,
  y = covariate,
  xmin = `0.025quant`,
  xmax = `0.975quant`,
  color = model
)) +
  geom_point(position = position_dodge(width = 0.7), size = 2) +
  geom_errorbar(position = position_dodge(width = 0.7), width = 0.25) +
  scale_color_manual(values = pal_modelos) +
  labs(
    x = "Posterior mean E[β | data]",
    y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.direction = "horizontal",
    legend.justification = "left"
  )

# ── 3. Posterior densities ───────────────────────────────────────────────────
marginals_beta <- fit_besag_sep$marginals.fixed

densities_df <- purrr::imap_dfr(marginals_beta, ~{
  as.data.frame(.x) |>
    tibble::as_tibble() |>
    dplyr::rename(beta = x, density = y) |>
    dplyr::mutate(covariate = covariate_labels[.y])
})

p_densities <- ggplot(densities_df, aes(x = beta, y = density)) +
  geom_line(color = pal_qualitat[1], linewidth = 0.6) +
  geom_vline(xintercept = 0, color = pal_qualitat[4], linetype = "dotted", linewidth = 0.8) +
  facet_wrap(~ covariate, scales = "free", ncol = 3) +
  labs(
    x = "β value",
    y = "Posterior density"
  ) +
  theme_minimal(base_size = 13)

# ── 4. Combine and export ────────────────────────────────────────────────────
mosaic_betas <- p_forest + p_densities + plot_layout(widths = c(1, 1.5))

ggsave(
  here::here("Figures", "English","Figure 6.eps"),
  plot   = mosaic_betas,
  width  = 13,
  height = 6.5,
  dpi    = 350,
  device = cairo_ps
)

mosaic_betas



```

## Figure 7. Spatial effects

Figure 7. Mean and uncertainty of structured (uᵢ) and unstructured (vᵢ) spatial effects.  
Top row: posterior means. Bottom row: posterior SDs indicating model confidence.


```{r Figure_7}


## ── Figura 7: Efectos espaciales estructurados (uᵢ) y no estructurados (vᵢ) ──

# 1. Extraer efectos aleatorios desde el modelo BYM separado
ran_u <- fit_besag_sep$summary.random$id_poly_u
ran_v <- fit_besag_sep$summary.random$id_poly_v

# 2. Agregar columnas al objeto sf si no existen
if (!"u_mean" %in% names(modelo_sf)) modelo_sf$u_mean <- NA_real_
if (!"u_sd"   %in% names(modelo_sf)) modelo_sf$u_sd   <- NA_real_
if (!"v_mean" %in% names(modelo_sf)) modelo_sf$v_mean <- NA_real_
if (!"v_sd"   %in% names(modelo_sf)) modelo_sf$v_sd   <- NA_real_

# 3. Asignar valores para polígonos válidos
modelo_sf$u_mean[keep_log] <- ran_u$mean
modelo_sf$u_sd  [keep_log] <- ran_u$sd
modelo_sf$v_mean[keep_log] <- ran_v$mean
modelo_sf$v_sd  [keep_log] <- ran_v$sd

# 4. Paleta continua para medias
pal_continuo <- paletteer::paletteer_c("grDevices::Zissou 1", 100)

# 5. Parámetros de leyenda
my_guide <- guide_colourbar(
  barwidth      = unit(0.3, "cm"),
  barheight     = unit(3, "cm"),
  direction     = "vertical",
  title.position = "top"
)

# 6. Mapas de media posterior y desviación estándar
p_u_mean <- ggplot(modelo_sf) +
  geom_sf(aes(fill = u_mean), color = NA) +
  scale_fill_gradientn(colours = pal_continuo, guide = my_guide, name = "mean(u)") +
  theme_minimal(base_size = 13) +
  theme(legend.justification = c("right", "top"))

p_u_sd <- ggplot(modelo_sf) +
  geom_sf(aes(fill = u_sd), color = NA) +
  scale_fill_viridis_c(option = "magma", direction = -1, guide = my_guide, name = "sd(u)") +
  theme_minimal(base_size = 13) +
  theme(legend.justification = c("right", "top"))

p_v_mean <- ggplot(modelo_sf) +
  geom_sf(aes(fill = v_mean), color = NA) +
  scale_fill_gradientn(colours = pal_continuo, guide = my_guide, name = "mean(v)") +
  theme_minimal(base_size = 13) +
  theme(legend.justification = c("right", "top"))

p_v_sd <- ggplot(modelo_sf) +
  geom_sf(aes(fill = v_sd), color = NA) +
  scale_fill_viridis_c(option = "magma", direction = -1, guide = my_guide, name = "sd(v)") +
  theme_minimal(base_size = 13) +
  theme(legend.justification = c("right", "top"))

# 7. Composición visual y exportación
p_ui_vi <- (p_u_mean + p_v_mean) / (p_u_sd + p_v_sd)

ggsave(
  filename = here::here("Figures", "Figure 7.eps"),
  plot     = p_ui_vi,
  width    = 10,
  height   = 10,
  dpi      = 350,
  device   = cairo_ps
)


```

